Online supplement accompanying `The Locally Gaussian Partial Correlation'

### 2020-09-16 lgpc_JBES_supp.tex
New set of comments from Dag

We proceed now to establish the notion of conditional independence between \(\Z_1\) and \(\Z_2\) using the LGPC in the multivariate case. The conditional densities of \(\Z_1|\Z_3\) and \(\Z_2|\Z_3\) in terms of the local Gaussian parameters are given by

@@

\begin{equation}
\fSigma_{1|3} = \fSigma_{11}-\fSigma_{13}\fSigma_{33}^{-1}\fSigma_{31}, \,\, \fSigma_{2|3} = \fSigma_{22}-\fSigma_{23}\fSigma_{33}^{-1}\fSigma_{32}.
\label{eq:localsigma}
\end{equation}
Under conditional independence between \(\Z_1\) and \(\Z_2\) given \(\Z_3\), the joint conditional density of \((\Z_1, \Z_2|\Z_3)\) is given by

@@

Observe, finally, that a more general pairwise scheme can be devised, as treated in detail by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}), Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) and Jordanger and Tjøstheim (\protect\hyperlink{ref-jordanger2017nonlinear}{2020}):

@@

We must next proceed by introducing some new concepts before stating the equivalents of assumptions A1 -- A2 by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}). These assumptions are similar but distinct from those used in Theorem 3.1 in the main article and Theorem 2 in Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}). They will also be used to prove the consistency result in Theorem \ref{thm:A1} at the end of this section, and they will be part of the assumptions used to prove the validity of the bootstrap in Section \ref{validity-bootstrap}.

@@

The conditions A1-- A2 are quite mild and discussed by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}). Note that they have been formulated for a scalar parameter \(\rho = \rho_{ij}\). To have them fulfilled for all possible combinations of \(i,j\), the assumption A1, for example can be easily formulated for the \(\{\Z_t\}\) process. Consistency for the local parameter estimates is required for consistency of the bootstrap:

@@

The proof of Theorem \ref{thm:A1} is mostly identical to the proof of Theorem 2.1 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) with trivial modifications to take into account the scalar nature of the parameter \(\rho_{ij}\) in the present case. A distributional limit theorem analogous to Theorem 2.2 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) can also be rephrased and proved as in that paper, but we refrain from doing this given the distributional results already stated in Section 3 of the main part of the paper.

@@

For fixed bandwidths we have from standard arguments, such as those provided by Hjort and Jones (\protect\hyperlink{ref-hjort1996locally}{1996}), that
\begin{equation}
\sqrt{nb^3}\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(0, \Jb^{-1}\Mb(\Jb^{-1})^T),
\label{eq:fixed-h}
\end{equation}
which corresponds to the usual rate for nonparametric density estimation. As \(\hh\rightarrow0\), however, we must take extra care when considering the asymptotic covariance matrix. Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}), Section 4, write the Taylor expansions of \(\Mb\) and \(\Jb\), now being functionals of three variables \((x_1, x_2, x_3)\) or \((z_1,z_2,z_3)\), as
\[\Mb = I_{\bm{M}} + II_{\bm{M}} + o(b^2) \,\,\, \textrm{ and } \Jb = I_{\bm{J}} + II_{\bm{J}} + o(b^2),\]
and consider each term. For example, we write their first term in the expansion of \(\Mb\) in the three-variate case as follows:
\[I_{\bm{M}} \sim \int K^2(w_1, w_2, w_3)Ab_wb_w^TA^Tf(\z + b_1w_1 + b_2w_2 + b_3w_3)\,\textrm{d}w_1\,\textrm{d}w_2\,\textrm{d}w_3,\]
where \(K\) is a product kernel, and, in our case, \(b_w\) is the vector defined by \(b_w^T = \begin{pmatrix} 1 & b_1w_1 & b_2w_2 & b_3w_3 \end{pmatrix}\), and \(A\) is the \(3\times 4\) matrix \(A = \begin{pmatrix} v & v_{z_1} & v_{z_2} & v_{z_3}\end{pmatrix}\) with \(\fv^T = \begin{pmatrix} v_1 & v_2 & v_3 \end{pmatrix}\), \(v_i = \partial L_n/\partial \rho_i\), where the index \(i=1,2,3\) represents the three local correlations, and \(v_{z_i} = \partial \fv/\partial z_i\). In the next step, they compute the matrix

\noindent \(\int K^2(w_1, w_2, w_3)b_wb_w^T\,\textrm{d}w_1\,\textrm{d}w_2\,\textrm{d}w_3\), which, by omitting all constant factors and exploiting that \(\int w^k K^2(w)\,\textrm{d}w = 0\) for \(k = 1,3\), in our case becomes the diagonal matrix
\[H = \begin{pmatrix} 1 & 0 & 0 & 0 \\ 0 & b_1^2w_1^2 & 0 & 0 \\ 0 & 0 & b_2^2w_2^2 & 0 \\ 0 & 0 & 0 & b_3^2w_3^2 \end{pmatrix}.\]

@@

We can use the bootstrap theory by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}, \protect\hyperlink{ref-lacal2018estimating}{2018}) more or less directly, which in turn is based on Gon\c{c}alves and White (\protect\hyperlink{ref-gonccalves2002bootstrap}{2002}, \protect\hyperlink{ref-gonccalves2004maximum}{2004}), but we spell out some details below. First, we need to specify the analogue of the conditions A1--A5 by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}), where we adapt their notation to our situation by replacing the parameter \(\theta\) with \(\rho_{ij}\), and the time series \(\{X_t, Y_t\}\) by our multivariate observations \(\{\Z_t\}\). Since \(\rho = \rho_{ij}\) is estimated using the two corresponding data vectors only, we can, for the time being and without loss of generality, assume that the observation series is bivariate: \({\Z_t} = \{Z_{i,t}, Z_{j,t}\}\). Note further that it suffices to consider the true pseudo-observations \(\{\Z_t\}\) instead of the empirical pseudo observations \(\{\hZ_t\}\) in the following because the difference between them is asymptotically negligible according to Theorem 3.2 in the main article.

To state the bootstrap results we need a few more concepts and three additional conditions. We block-bootstrap the entire vector series \(\{\Z_t\}\). This automatically leads to the same block-bootstrap for any pair of components which we have denoted by \(\{Z_{i,t}, Z_{j,t}\}\) above. Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) use both the block bootstrap as well as the stationary bootstrap, and the results below hold irrespective of this particular choice. We will therefore not distinguish between them here, but rather refer the reader to Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) for further details on this distinction. Bootstrapped quantities will be denoted with a star (\(*\)) in the following.

@@

Assumption A4 sets the stage for the block bootstrapping, whereas A5 is needed in proving the validity of the bootstrap central limit theorem. The assumption A4 is not restrictive, and again we refer to Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) for more discussion.

@@

Theorems \ref{thm:E1} and \ref{thm:E2} are formulated in their simplest case here, which concerns only one local parameter. A more complicated version is given by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) concerning the five parameter case. The proofs are essentially the same and are spelled out in some detail in the supplementary material of that paper as Theorems D1 and D2.

@@

\emph{Proof:} The proof of Theorem \ref{thm:E2} is given as the proof of Theorem D2 in the online appendix accompanying Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}). This proof can be adapted to prove Corollary \ref{cor:E4}. The starting point consists in replacing \(L_{n,b}^*\) and \(L_{n,b}\) by \(L^* = \sum_{i,j} a_{ij}L_{ij,n,b}^*\) and \(L = \sum_{i,j}a_{ij}L_{ij,n,b}\), respectively, where the sum runs over all possible pairs \((i,j)\) corresponding to \(\rho_{ij}^{1,2}\), \(\rho_{ij}^{1,3}\), \(\rho_{ij}^{2,3}\) that are included in the \(\frho\)-vector. The derivations in the proof of Theorem D2 by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) deal with the more complicated five-parameter case. In the present case, however, the vector \(\frho\) consists of correlations \(\rho_{ij}\) that have been estimated using the one-parameter pairwise setup, where, cf.~Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}), the quantities \(J_{ij}\) and \(M_{ij}\) are scalars. Here \(L_{ij,n.b}\) is the log likelihood for the pair \((i,j)\) treated in the proof of Theorem D2, and \(a_{ij}\) is a corresponding sequence of real numbers. Because of the linearity of the expressions for \(L^*\) and \(L\) it is easily checked by going step by step that the proof of Theorem D2 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) can be carried through for \(L^*\) and \(L\). The only difference is that the quantities \(J_{n,b}\) and \(M_{n,b}\) now have to be replaced by \(J_{n.b}(a) = \E(-\sum_{i,j}a_{ij}\nabla^2 L_{ij,n,b}(\rho_{0,b}))\) and \(\Var(\sqrt{nb_1b_2}\sum_{i,j} a_{ij}\nabla L_{ij,n,b})\). This results in joint asymptotic bootstrap normality using the Cramer-Wold device. Applying the delta rule now proves the Corollary. \(\blacksquare\)

Let \(\M_{ij}\) and \(\J_{ij}\) be the \(M\)-matrix and \(J\)-matrix appearing in the proof of Corollary \ref{cor:E4} for a fixed \(\rho_{ij}\). Note that in the one-parameter case of \(\rho_{ij}\) in Corollary \ref{cor:E4}, to first order the covariance matrix \(\fLambda\) is a diagonal matrix having ordered values of \(M_{ij}/J_{ij}^2\) along the diagonal. This can be seen using the methods of proofs of Theorem 2 of Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}) and Theorem 2 Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}). Note also that the 3-coordinate case of alternative (i) can be stated and proved in the same way, but a different scaling factor is required to avoid a singular covariance matrix as seen in the proof of Theorem 3.1 in the main article. The 3-coordinate case is in fact closer to the situation in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}).

@@

Consider the general case \(\Z = (\Z_1, \Z_2, \Z_3)\) where each component is \(d_i\)-dimensional correspondingly, with \(i = 1,2,3\), and \(\Z\), naturally, is \(d_1+d_2+d_3\)-dimensional, and consider also the local partial correlation matrix of \((\Z_1,\Z_2|\Z_3)\). In alternative (iii), where \(d_3 = 1\), in Section \ref{chap:estimation} this matrix is given by equation \eqref{eq:offdiag3} with
\[
\alpha_{ij} = \rho(Z_{1,i},Z_{2,j}|Z_3), \;\; i=1,\ldots,d_1; j = 1,\dots,d_2.
\]
This forms a \(d_1 \times d_2\) matrix \({\falpha}\). One can test for conditional independence between \(Z_{1,i}\) and \(Z_{2,j}\) given \(\Z_3\) by testing whether \(\alpha_{ij}=0\), or one can test for conditional independence between \(\Z_1\) and \(\Z_2\) given \(\Z_3\) by testing whether \({\falpha} = \{\alpha_{ij}\} = {\bm{0}}\). The latter case is in a sense the most general, and we will stick to that in the sequel. One can then test for \({\falpha} = {\bm{0}}\) by using a matrix norm, for instance the Frobenius norm, where \(||{\falpha}|| = \sqrt{\sum_{i,j} \alpha_{ij}^2} = 0\). A more convenient test statistic is \(\sum_{i,j} \alpha_{ij}^2\). To follow the notation of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) we write, letting \(\z=(\z_1,\z_2,\z_3)\),
\[
T = \int_S h(\frho(\z)) \di F(\z)
\]
where \(S\) is an a area of integration, \(F\) is the cumulative distribution function of \(\Z\), and \(\frho = \frho(\z)\), as before, is a vector consisting of all possible pairs that can be formed by picking pairwise components from \(\Z\). The number of components of this vector depends on which of the alternatives models (i)--(v) in Section \ref{chap:estimation} one considers. It is reasonable to express \(T\) as a function of \(\frho\) since all partial correlations can be expressed as a function of this vector (see e.g.~again equation \eqref{eq:offdiag3}). One possible choice of the function \(h\) is \(h(\frho(\z)) = \sum_{i,j} \alpha_{i,j}^2(z_{1,i},z_{2,j}|\ztwo)\), which is a function of \(\frho\) since the \(\alpha\)-s are functions of \(\frho\). The corresponding function in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) is \(h(\theta(x))\), where \(\theta(x)\) is at most 5-dimensional and \(\x\) is bivariate. As in that paper we proceed in two stages. First we let the bandwidth vector \(\bb\) be fixed and let \(n \to \infty\), and then we let \(\bb = (b_1,b_2) = (b,b) \to \bm{0}\).
Accordingly we define
\[
T_b = \int_S h(\frho_b(\z)) \di F(\z) 
\]
where \(\frho_b\) is the population quantity for a fixed bandwidth \(\bb\), and \(T_{n,b}\) by
\[
T_{n,b} = \int_S h(\frho_{b,n}(\z)) \di F_n(\z),
\]
where \(\frho_{b,n}(\z)\) is the estimate of \(\frho(\z)\) obtained by pairwise likelihood for each component of \(\frho(\z)\).

@@

It is more difficult to obtain the limiting distribution. We will follow the technique in the proof of Theorem 3.1 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}). See also Joe (\protect\hyperlink{ref-joe1989estimation}{1989}). The details are a bit more involved in the present case because of the larger dimension and because we need to involve the pairwise log likelihood in the procedure.

@@

\begin{thm} 
Let the assumptions A, D--F of Theorem 3 of Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}) and the conditions of Theorem 3.2 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) (which includes conditions A1 and A2 of this supplement) be fulfilled. Then,
as $n \to \infty$,
$$
\sqrt{n}\big(C_{n}(A_b)\big)^{-\frac{1}{2}}(T_{n,b} - T_b) \buildrel {d} \over {\to} {\cal N}(0,1),
$$
where
\begin{align*}
A_b(\z)  & = \int_{S} \left[\sum_{i,j}\left\{\frac{\partial}{\partial \rho_{ij}} h(\frho_b,(\fv)\right\}J_{ij}^{-1}(\rho_{ij,b}(\fv_{ij}))K_b(\z_{ij}- \fv_{ij})u_{ij}(\z_{ij},\rho_{ij,b}(\fv_{ij}))\right] \di F(\fv) \\
& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad + h(\frho_b(\z))1_S(\z)
\end{align*}
and 
\begin{align}
C_n(A_b) &= \int A_b^2(\z) \di F(\z) + \frac{1}{n}\sum_{s \neq t}\int A_b(\z)A_b(\fu) \di F^{s,t}(\z,\fu) \nonumber \\
& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad - n\int A_b(\z)A_b(\fu) \di F(\z) \di F(\fu)
\label{eq:t0}
\end{align}
with $J_{ij}$ being the bivariate version of $\J_b$ defined in Theorem 3.1, and, with $F^{s,t}(\z,\fu)$ being the joint cumulative distribution function of $(\Z_s,\Z_t)$, and where $\{\Z_{t}, t=1,\ldots,n\}$ are the $n$ observations of $\Z$ and $u_{ij}(\Z_{ij}, \rho_{ij,b}(\fv_{ij})) = \frac{\partial}{\partial \rho_{ij,b}}\log\psi(\Z_{ij},\rho_{ij,b}(\fv_{ij}))$.

@@

Note that the sum in the second term in the definition of \(\fC_n(\fA_b)\) taken together with the third term converges due to the geometric \(\alpha\)-mixing condition of A1. It is also important to note that conditional independence is not assumed in Theorem \ref{thm:F2}. This will be exploited in Section \ref{chap:pitman}. See also Theorem 3.2 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) and its relation to Corollary 3.1 in that paper.

>>>>

We proceed now to establish the notion of conditional independence between \(\Z_1\) and \(\Z_2\) using the LGPC in the multivariate case. The conditional densities of \(\Z_1|\Z_3\) and \(\Z_2|\Z_3\) in terms of the local Gaussian parameters at the point \((\z_1, \z_2, \z_3)\) are given by

@@

\begin{equation}
\fSigma_{1|3} = \fSigma_{11}-\fSigma_{13}\fSigma_{33}^{-1}\fSigma_{31}, \,\, \fSigma_{2|3} = \fSigma_{22}-\fSigma_{23}\fSigma_{33}^{-1}\fSigma_{32}.
\label{eq:localsigma}
\end{equation}
Under conditional independence between \(\Z_1\) and \(\Z_2\) given \(\Z_3\), the joint conditional density of \((\Z_1, \Z_2|\Z_3)\) at the point \((\z_1, \z_2, \z_3)\) is given by

@@

Observe, finally, that a more general pairwise scheme can be devised, as treated in detail by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}), Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) and Jordanger and Tjøstheim (\protect\hyperlink{ref-jordanger2017nonlinear}{2020}):

@@

We must next proceed by introducing some new concepts before stating the equivalents of assumptions A1 -- A2 by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}). These assumptions are similar but distinct from those used in Theorem 3.1 in the main article and Theorem 2 in Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}). They will also be used to prove the consistency result in Theorem \ref{thm:A1} at the end of this section, and they will be part of the assumptions used to prove the validity of the bootstrap in Section \ref{validity-bootstrap}.

@@

The conditions A1-- A2 are quite mild and discussed by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}). Note that they have been formulated for a scalar parameter \(\rho = \rho_{ij}\). To have them fulfilled for all possible combinations of \(i,j\), the assumption A1, for example can be easily formulated for the \(\{\Z_t\}\) process. Consistency for the local parameter estimates is required for consistency of the bootstrap:

@@

The proof of Theorem \ref{thm:A1} is mostly identical to the proof of Theorem 2.1 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) with trivial modifications to take into account the scalar nature of the parameter \(\rho_{ij}\) in the present case. A distributional limit theorem analogous to Theorem 2.2 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) can also be rephrased and proved as in that paper, but we refrain from doing this given the distributional results already stated in Section 3 of the main part of the paper.

@@

For fixed bandwidths we have from standard arguments, such as those provided by Hjort and Jones (\protect\hyperlink{ref-hjort1996locally}{1996}), that
\begin{equation}
\sqrt{nb^3}\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(0, \Jb^{-1}\Mb(\Jb^{-1})^T),
\label{eq:fixed-h}
\end{equation}
which corresponds to the usual rate for nonparametric density estimation. As \(\hh\rightarrow0\), however, we must take extra care when considering the asymptotic covariance matrix. Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}), Section 4, write the Taylor expansions of \(\Mb\) and \(\Jb\), now being functionals of three variables \((x_1, x_2, x_3)\) or \((z_1,z_2,z_3)\), as
\[\Mb = I_{\bm{M}} + II_{\bm{M}} + o(b^2) \,\,\, \textrm{ and } \Jb = I_{\bm{J}} + II_{\bm{J}} + o(b^2),\]
and consider each term. For example, we write their first term in the expansion of \(\Mb\) in the three-variate case as follows:
\[I_{\bm{M}} \sim \int K^2(w_1, w_2, w_3)Ab_wb_w^TA^Tf(\z + b_1w_1 + b_2w_2 + b_3w_3)\,\textrm{d}w_1\,\textrm{d}w_2\,\textrm{d}w_3,\]
where \(K\) is a product kernel, and, in our case, \(b_w\) is the vector defined by \(b_w^T = \begin{pmatrix} 1 & b_1w_1 & b_2w_2 & b_3w_3 \end{pmatrix}\), and \(A\) is the \(3\times 4\) matrix \(A = \begin{pmatrix} v & v_{z_1} & v_{z_2} & v_{z_3}\end{pmatrix}\) with \(\fv^T = \begin{pmatrix} v_1 & v_2 & v_3 \end{pmatrix}\), \(v_i = \partial L_n/\partial \rho_i\), where the index \(i=1,2,3\) represents the three local correlations, and \(v_{z_i} = \partial \fv/\partial z_i\). In the next step, they Taylor expand \(f\) and compute the matrix \(\int K^2(w_1, w_2, w_3)b_wb_w^T\,\textrm{d}w_1\,\textrm{d}w_2\,\textrm{d}w_3\), which, by omitting all constant factors and exploiting that \(\int w^k K^2(w)\,\textrm{d}w = 0\) for \(k = 1,3\), in our case becomes the diagonal matrix
\[H = \begin{pmatrix} 1 & 0 & 0 & 0 \\ 0 & b_1^2w_1^2 & 0 & 0 \\ 0 & 0 & b_2^2w_2^2 & 0 \\ 0 & 0 & 0 & b_3^2w_3^2 \end{pmatrix}.\]

@@

We can use the bootstrap theory by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}, \protect\hyperlink{ref-lacal2018estimating}{2019}) more or less directly, which in turn is based on Gon\c{c}alves and White (\protect\hyperlink{ref-gonccalves2002bootstrap}{2002}, \protect\hyperlink{ref-gonccalves2004maximum}{2004}), but we spell out some details below. First, we need to specify the analogue of the conditions A1--A5 by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}), where we adapt their notation to our situation by replacing the parameter \(\theta\) with \(\rho_{ij}\), and the time series \(\{X_t, Y_t\}\) by our multivariate observations \(\{\Z_t\}\). Since \(\rho = \rho_{ij}\) is estimated using the two corresponding data vectors only, we can, for the time being and without loss of generality, assume that the observation series is bivariate: \({\Z_t} = \{Z_{i,t}, Z_{j,t}\}\). Note further that it suffices to consider the true pseudo-observations \(\{\Z_t\}\) instead of the empirical pseudo observations \(\{\hZ_t\}\) in the following because the difference between them is asymptotically negligible according to Theorem 3.2 in the main article.

To state the bootstrap results we need a few more concepts and three additional conditions. We block-bootstrap the entire vector series \(\{\Z_t\}\). This automatically leads to the same block-bootstrap for any pair of components which we have denoted by \(\{Z_{i,t}, Z_{j,t}\}\) above. Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) use both the block bootstrap as well as the stationary bootstrap, and the results below hold irrespective of this particular choice. We will therefore not distinguish between them here, but rather refer the reader to Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) for further details on this distinction. Bootstrapped quantities will be denoted with a star (\(*\)) in the following.

@@

Assumption A4 sets the stage for the block bootstrapping, whereas A5 is needed in proving the validity of the bootstrap central limit theorem. The assumption A4 is not restrictive, and again we refer to Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) for more discussion.

@@

Theorems \ref{thm:E1} and \ref{thm:E2} are formulated in their simplest case here, which concerns only one local parameter. A more complicated version is given by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) concerning the five parameter case. The proofs are essentially the same and are spelled out in some detail in the supplementary material of that paper as the proofs of Theorems D1 and D2.

@@

\emph{Proof:} The proof of Theorem \ref{thm:E2} is given as the proof of Theorem D2 in the online appendix accompanying Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}). This proof can be adapted to prove Corollary \ref{cor:E4}. The starting point consists in replacing \(L_{n,b}^*\) and \(L_{n,b}\) by \(L^* = \sum_{i,j} a_{ij}L_{ij,n,b}^*\) and \(L = \sum_{i,j}a_{ij}L_{ij,n,b}\), respectively, where the sum runs over all possible pairs \((i,j)\) corresponding to \(\rho_{ij}^{1,2}\), \(\rho_{ij}^{1,3}\), \(\rho_{ij}^{2,3}\) that are included in the \(\frho\)-vector. The derivations in the proof of Theorem D2 by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) deal with the more complicated five-parameter case. In the present case, however, the vector \(\frho\) consists of correlations \(\rho_{ij}\) that have been estimated using the one-parameter pairwise setup, where, cf.~Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}), the quantities \(J_{ij}\) and \(M_{ij}\) are scalars. Here \(L_{ij,n.b}\) is the log likelihood for the pair \((i,j)\) treated in the proof of Theorem D2, and \(a_{ij}\) is a corresponding sequence of real numbers. Because of the linearity of the expressions for \(L^*\) and \(L\) it is easily checked by going step by step that the proof of Theorem D2 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) can be carried through for \(L^*\) and \(L\). The only difference is that the quantities \(J_{n,b}\) and \(M_{n,b}\) now have to be replaced by \(J_{n.b}(a) = \E(-\sum_{i,j}a_{ij}\nabla^2 L_{ij,n,b}(\rho_{0,b}))\) and \(\Var(\sqrt{nb_1b_2}\sum_{i,j} a_{ij}\nabla L_{ij,n,b})\). This results in joint asymptotic bootstrap normality using the Cramer-Wold device. Applying the delta rule now proves the Corollary. \(\blacksquare\)

Let \(\M_{ij}\) and \(\J_{ij}\) be as in the proof of Corollary \ref{cor:E4} for a fixed \(\rho_{ij}\). Note that in the one-parameter case of \(\rho_{ij}\) in Corollary \ref{cor:E4}, to first order the covariance matrix \(\fLambda\) is a diagonal matrix having ordered values of \(M_{ij}/J_{ij}^2\) along the diagonal. This can be seen using the methods of proofs of Theorem 2 of Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}) and Theorem 2 Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}). Note also that the 3-coordinate case of alternative (i) can be stated and proved in the same way, but a different scaling factor is required to avoid a singular covariance matrix as seen in the proof of Theorem 3.1 in the main article. The 3-coordinate case is in fact closer to the situation in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}).

@@

Consider the general case \(\Z = (\Z_1, \Z_2, \Z_3)\) where each component is \(d_i\)-dimensional correspondingly, with \(i = 1,2,3\), and \(\Z\), naturally, is \(d_1+d_2+d_3\)-dimensional, and consider also the local partial correlation matrix of \((\Z_1,\Z_2|\Z_3)\). In alternative (iii), where \(d_3 = 1\), in Section \ref{chap:estimation} this matrix is given by equation \eqref{eq:offdiag3} with
\[
\alpha_{ij} = \rho(Z_{1,i},Z_{2,j}|Z_3), \;\; i=1,\ldots,d_1; j = 1,\dots,d_2.
\]
This forms a \(d_1 \times d_2\) matrix \({\falpha}\). One can test for conditional independence between \(Z_{1,i}\) and \(Z_{2,j}\) given \(\Z_3\) by testing whether \(\alpha_{ij}=0\), or one can test for conditional independence between \(\Z_1\) and \(\Z_2\) given \(\Z_3\) by testing whether \({\falpha} = \{\alpha_{ij}\} = {\bm{0}}\). The latter case is in a sense the most general, and we will stick to that in the sequel. One can then test for \({\falpha} = {\bm{0}}\) by using a matrix norm, for instance the Frobenius norm, where \(||{\falpha}|| = \sqrt{\sum_{i,j} \alpha_{ij}^2} = 0\). A more convenient test statistic is \(\sum_{i,j} \alpha_{ij}^2\). To follow the notation of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) we write, letting \(\z=(\z_1,\z_2,\z_3)\),
\[
T = \int_S h(\frho(\z)) \di F(\z)
\]
where \(S\) is an a area of integration, \(F\) is the cumulative distribution function of \(\Z\), and \(\frho = \frho(\z)\), as before, is a vector consisting of all possible pairs that can be formed by picking pairwise components from \(\Z\). The number of components of this vector depends on which of the alternatives models (i)--(v) in Section \ref{chap:estimation} one considers. It is reasonable to express \(T\) as a function of \(\frho\) since all partial correlations can be expressed as a function of this vector (see e.g.~again equation \eqref{eq:offdiag3}). One possible choice of the function \(h\) is \(h(\frho(\z)) = \sum_{i,j} \alpha_{i,j}^2(z_{1,i},z_{2,j}|\ztwo)\), which is a function of \(\frho\) since the \(\alpha\)-s are functions of \(\frho\). The corresponding function in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) is \(h(\theta(\x), \x)\), where \(\theta(\x)\) is at most 5-dimensional and \(\x\) is bivariate. As in that paper we proceed in two stages. First we let the bandwidth vector \(\bb\) be fixed and let \(n \to \infty\), and then we let \(\bb = (b_1,b_2) = (b,b) \to \bm{0}\).
Accordingly we define
\[
T_b = \int_S h(\frho_b(\z)) \di F(\z) 
\]
where \(\frho_b\) is the population quantity for a fixed bandwidth \(\bb\), and \(T_{n,b}\) by
\[
T_{n,b} = \int_S h(\frho_{b,n}(\z)) \di F_n(\z),
\]
where \(\frho_{b,n}(\z)\) is the estimate of \(\frho(\z)\) obtained by pairwise likelihood for each component of \(\frho(\z)\).

@@

It is more difficult to obtain the limiting distribution. We will follow the technique in the proof of Theorem 3.2 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}). See also Joe (\protect\hyperlink{ref-joe1989estimation}{1989}). The details are a bit more involved in the present case because of the larger dimension and because we need to involve the pairwise log likelihood in the procedure.

@@

\begin{thm} 
Let the assumptions A, D--F of Theorem 3 of Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}) and the conditions of Theorem 3.2 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) (which includes conditions A1 and A2 of this supplement) be fulfilled. Then,
as $n \to \infty$,
$$
\sqrt{n}\big(C_{n}(A_b)\big)^{-\frac{1}{2}}(T_{n,b} - T_b) \buildrel {d} \over {\to} {\cal N}(0,1),
$$
where
\begin{align*}
A_b(\z)  & = \int_{S} \left[\sum_{i,j}\left\{\frac{\partial}{\partial \rho_{ij}} h(\frho_b,(\fv))\right\}J_{ij}^{-1}(\rho_{ij,b}(\fv_{ij}))K_b(\z_{ij}- \fv_{ij})u_{ij}(\z_{ij},\rho_{ij,b}(\fv_{ij}))\right] \di F(\fv) \\
& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad + h(\frho_b(\z))1_S(\z)
\end{align*}
and 
\begin{align}
C_n(A_b) &= \int A_b^2(\z) \di F(\z) + \frac{1}{n}\sum_{s \neq t}\int A_b(\z)A_b(\fu) \di F^{s,t}(\z,\fu) \nonumber \\
& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad - n\int A_b(\z)A_b(\fu) \di F(\z) \di F(\fu)
\label{eq:t0}
\end{align}
with $J_{ij}$ being the bivariate version of $\J_b$ defined in Theorem 3.1, and, with $F^{s,t}(\z,\fu)$ being the joint cumulative distribution function of $(\Z_s,\Z_t)$, and where $\{\Z_{t}, t=1,\ldots,n\}$ are the $n$ observations of $\Z$ and $u_{ij}(\Z_{ij}, \rho_{ij,b}(\fv_{ij})) = \frac{\partial}{\partial \rho_{ij,b}}\log\psi(\Z_{ij},\rho_{ij,b}(\fv_{ij}))$.

@@

Note that the sum in the second term in the definition of \(\fC_n(\fA_b)\) taken together with the third term converges due to the geometric \(\alpha\)-mixing condition of A1. It is also important to note that conditional independence is not assumed in Theorem \ref{thm:F2}. This will be exploited in Section \ref{chap:pitman}. See also Theorem 3.2 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) and its relation to Corollary 3.1 in that paper.
